Asymptotic Nonlinear Solution for Electrokinetic Flow Around an Ion-Exchange Spherical Particle

### 2012-09-21 notes/work/cubic_report.tex
Add cubic report

\documentclass[10pt]{ijnam}
\hsize=5.5 true in
\textheight=8.4 true in
\topmargin 1in

%% information for the editor
\def\currentvolume{1}
\def\currentissue{1}
\def\currentyear{2004}
\def\month{October}
\pagespan{1}{18}
\copyrightinfo{2012}{} % copyright year

@@

\title{Asymptotic and Numerical Solution for Electrokinetic Flow Around a Spherical Particle}

% author and address
\author{Roman Zeyde \and Irad Yavneh}
\address{Department of Computer Science, Technion -- Israel Institute of Technology, Haifa 32000, Israel}
\email{romanz@cs.technion.ac.il \and irad@cs.technion.ac.il}

% two authors have the same address
%\author{Ory Schnitzer \and Ehud Yariv}
%\address{Department of Mathematics, Technion -- Israel Institute of Technology, Haifa 32000, Israel}
%\email{oschnitt@gmail.com \and udi@math.technion.ac.il}

% communicated
%\commby{Hilbert}

% Received by the editors ?
\date{DATE}

% \subjclass[2000]{35R35, 49J40, 60G40}

\abstract{
A numerical framework for the simulation of electrokinetic 
migration of particles in an electrolyte solution 
due to the application of an external electric field is presented.
The electrokinetic transport process is
described by a system of nonlinear partial differential equations (PDE).
A thin boundary layer forms around the particle due to strong electrostatic forces. The resulting
scale disparity of the boundary layer
is used to derive nonlinear effective boundary conditions for the
numerical solver
using the specific chemical properties of the particle.
The nonlinear system is discretized and an iterative Newton solver is
constructed automatically from the discrete equations and boundary 
conditions.
An asymptotic analytical solution is 
derived for the validation of the solver. 
Numerical results are obtained for an ion-exchanger and 
for a surface charged inert particle. The numerical results 
are compared to the asymptotic solutions, and good correspondence is achieved.}

\keywords{CFD, electrokinetics, asymptotics, ion-exchanger, electrophoresis}

@@

Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge,
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle
size, an analytical asymptotic solution for the Debye layer
dynamics can be derived \cite{yariv2010asymptotic}.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=0.35\textwidth]
            {figs/debye.eps}
        \caption{Schematic structure of the double layer:
        (1) charged particle surface, (2) Debye layer, (3) fluid bulk.
        If the particle surface has positive (black) surface charge,
        it attracts negative (white) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        The zeta potential is defined as the voltage drop across 
        the Debye layer (2).}
        \label{fig:EDL}
    \end{center}
\end{figure}
% \begin{tikzpicture}\end{tikzpicture}

The variables of the electrokinetic flow equations are the electrostatic
potential $\varphi$, fluid velocity $\bv$ and its pressure $p$, and
ionic concentration $c$.
The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.
The partial differential equations that describe the system dynamics
under an external electric field, are coupled and nonlinear, and
in general they have no analytic solution. 
Moreover, any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle size. 
A closed-form linear asymptotic solution has been developed for
spherical particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena emerge.
This interesting regime has not yet been explored extensively.
Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects in many nanotechnology applications,
nano-fluidic devices, packed bed separation, desalination processes and various 
electrophoresis applications \cite{howse2007self,paxton2004catalytic,pumera2010electrochemically,nischang2007,leinweber2004}.
Because the model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver for this system is of importance.

The goal of this work is to develop and implement a software toolbox for the generation
of iterative numerical solvers for nonlinear electrokinetic problems, and to apply it to the study
of systems that have no closed-form solutions, such as ion-exchanger migration
\cite{yariv2010migration} and the electrophoresis problem \cite{schnitzer2012surface}.
This solver can be used to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also to possibly lead to further analytical developments, based on new scaling behavior
discovered numerically. We may also target interesting physical
phenomena that have been observed experimentally.
One such phenomenon is nonlinear electrokinetic vortex flow, 
which is useful for microfluidic mixing applications \cite{wang2004mix, ben2002vortex},

The paper is structured as follows: the governing equations of the electrokinetic model
are derived in Section \ref{sec:equations}. The algorithm for the discretization and 
the solution of the system is described in Section \ref{sec:algorithm}.
An analytical asymptotic solution for the nonlinear system is derived 
and presented in Section \ref{sec:asymp}.
Numerical results and their comparison to an analytical asymptotic solution 
are presented Section \ref{sec:results} and discussed in Section \ref{sec:discussion}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations} \label{sec:equations}

\subsection{Nondimensionalization}
Spatial coordinates $\br$ are normalized by $a^*$ 
(the characteristic size of the particle).
Positive and negative ionic concentrations are denoted by $c_+$ and $c_-$, respectively, and
are normalized by the ambient ionic concentration $c^*$. The ionic valences are $\pm z$.
The dimensional ionic fluxes, $\bj_\pm$, are given by the Nernst-Plank equation 
(using the Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$, 
where $D^*$ is the ionic diffusivity):
\begin{equation}
\bj^*_\pm = 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{equation}

The electric potential $\varphi$ is normalized by the thermal voltage:
$\varphi^* = {k_B^* T^*}/{z e^*} = {R^* T^*}/{z F^*},$
where $R^* = k_B^* N_A$, $F^* = e^* N_A$ is the Faraday number and $N_A$ is the Avogadro constant.

The dimensional Stokes equations with electrostatic forces are given by:
\begin{equation} \begin{array}{cc}
\bnabla \cdot \bv = 0, &
-\bnabla p^* + \mu^* \bLaplacian \bv^* + \eps^* \Laplacian \varphi^* 
\bnabla \varphi^* = 0.
\end{array}\end{equation}

The velocity $\bv$ is normalized by $v^* = {\eps^* (\varphi^*)^2}/{a^* \mu^*}$,
and the pressure $p$ is normalized by $p^* = {\mu^* v^*}/{a^*} = {\eps^* (\varphi^*)^2}/{(a^*)^2}$.
The fluxes $\bj^*_\pm$ are normalized by $j^* = {D^* c^*}/{a^*}$.

The dimensional Poisson equation can be written as
\begin{equation}
(c^*_+ - c^*_-) z e^* = -\eps^* \Laplacian \varphi^*,
\end{equation}

After nondimensionalization, the Nernst-Plank, Stokes and Poisson equations read:
\begin{equation} \label{eq:nernst}
\bj_\pm = 
-\bnabla c_\pm + \alpha \bv c_\pm \mp c_\pm \bnabla \varphi,
\end{equation}
\begin{equation} \label{eq:stokes}
\begin{array}{cc}
\bnabla \cdot \bv = 0, &
-\bnabla p + \bLaplacian \bv + \Laplacian \varphi \bnabla \varphi = 0, 
\end{array}
\end{equation}
\begin{equation} \label{eq:poisson}
c_+ - c_- = -2\delta^2 \Laplacian \varphi,
\end{equation}
where $\delta = {\delta^*}/{a^*}$ is the dimensionless Debye thickness 
($\delta^* = \sqrt{{\eps^* \varphi^*}/{2 c^* e^* z}}$) and 
$\alpha = {v^* a^*}/{D^*} = {\eps^* (\varphi^*)^2}/{\mu^* D^*}$ 
is the dimensionless Peclet number.
The force on the particle is denoted by $\bF$ normalized by $\mu^* v^* a^* = \eps^* (\varphi^*)^2$.
The stress tensor is denoted by $\tT$ and normalized by $p^* = \mu^* v^* / a^*$.
The electric field is denoted by $\bE$ and normalized by $E^* = \varphi^* / a^*$.
The surface charge is denoted by $\sigma$ and normalized by $\sigma^* = \eps^* \varphi^* / \delta^*$.

The ionic fluxes are given by ion diffusion, electrostatic forces and ion advection by the fluid
in \eqref{eq:nernst}. Due to conservation of ion concentrations, the fluxes are divergence-free:
\begin{equation*}
\bnabla \cdot \bj_\pm = 0.
\end{equation*}
These equations may be re-written by using salt $c = (c_+ + c_-)/2$ and charge $q = (c_+ - c_-)/2$.
Thus, salt and charge fluxes are defined by:
\begin{equation}\label{eq:fluxes}\begin{array}{cc}
  \bj = \frac{\bj_+ + \bj_-}{2} = -\bnabla c - q \bnabla \varphi + \alpha \bv c, & 
  \bi = \frac{\bj_+ - \bj_-}{2} = -\bnabla q - c \bnabla \varphi + \alpha \bv q,
\end{array}\end{equation}
and are divergence-free as well:
\begin{equation}\label{eq:zero_flux}\begin{array}{cc}
\bnabla \cdot \bj = 0, & 
\bnabla \cdot \bi = 0. 
\end{array}\end{equation}
In addition, the charge concentration is governed by the Poisson equation \eqref{eq:poisson}
\begin{equation}
q = -\delta^2 \Laplacian \varphi.
\end{equation}

\subsection{Boundary conditions}
We assume that a constant electric field of nondimensional magnitude $\beta$,
is applied to the system, in the direction denoted by the unit vector $\ui$.
As a result, the particle will equilibrate at 
a non-zero steady-state velocity $\cU \ui$, which is
a function of the applied field, and is not known a priori.

We employ the reference frame of the particle.
Far away from the particle, we have a uniform applied electrostatic field, uniform flow,
and ambient ionic concentrations:
\begin{equation}\label{eq:bndcond_inf}\begin{array}{ccc}
\bv \rightarrow -\cU \ui, &
\bnabla \varphi \rightarrow -\beta\ui, &
 c_\pm \rightarrow 1.
\end{array}\end{equation}

We consider a spherical particle.
The chemical properties of the particle's surface $\mathcal{S}$ (defined at $r=1$ 
with normal $\bn = \brhat$), determine the boundary conditions on $\mathcal{S}$.
We consider the following scenarios:
\subsubsection{Ion-exchanger \cite{yariv2010migration}}
We assume high conductance of the particle (yielding a uniform electric potential),
zero slip, anion impermeability, cation selectivity 
(with fast Butler-Volmer\cite{bard2000book} kinetics
$\bn \cdot \bj_+ = k (1 - c_+/\gamma)$, where $k \rightarrow \infty$):
\begin{equation}
\begin{array}{cccc}
\varphi = \cV, &
\bv = \bzero, &
\bn \cdot \bj_- = 0, &
c_+ = \gamma,
\end{array}
\end{equation}
where $\gamma$ is the ratio between the cationic concentration on the surface and the cationic
concentration in the fluid bulk and $\cV$ is an unknown constant electric potential at
the ion-exchanger surface.

\subsubsection{Inert particle \cite{schnitzer2012surface}}
We assume constant surface charge
zero slip and ion impermeability:
\begin{equation}
\begin{array}{cccc}
\bn \cdot (\bnabla \varphi - \chi \bnabla \varphi_s) = \delta^{-1} \sigma, &
\bv = \bzero, &
\bn \cdot \bj_\pm = 0,
\end{array}
\end{equation}
where $\chi$ is the ratio between the dielectric constants of the 
particle and the fluid.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of scales}
When $\delta \ll 1$, 
the fluid remains electroneutral ($q \approx 0$, thus $c_+ \approx c_-$), 
except inside a thin boundary layer of width $O(\delta)$ 
surrounding the particle (the ``Debye layer'').
This boundary layer makes it very hard to construct 
a direct numerical solver for the full electrokinetic problem.
However, we exploit the scale disparity at the Debye layer, 
which allows the PDE system to be rewritten in the ``inner'' region 
as an ODE system using the asymptotic limit of $\delta \rightarrow 0$.
This system can be integrated and solved analytically,
yielding ``effective'' boundary conditions near
the particle surface for the ``outer'' bulk region equations.

\subsubsection  {Bulk-scale equations}
The Nernst-Planck equations above \eqref{eq:fluxes} for the fluid bulk outside
the boundary layer, using bulk variables denoted by uppercase letters, read:
\begin{equation} \begin{array}{cccc} \label{eq:bulk_flux}
  Q = 0, &
  C = C_+ = C_-, &
\bJ = -\bnabla C + \alpha \bV C, &
\bI = -C \bnabla \varPhi.
\end{array}\end{equation}
Because the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following 
salt and charge conservation equations -- the salt and charge fluxes above 
\eqref{eq:bulk_flux} are divergence-free
\eqref{eq:zero_flux}:
\begin{equation} \begin{array}{ccc}
\label{eq:salt_charge}
\Laplacian C - \alpha \bV \cdot \bnabla C = 0, &
\bnabla \cdot \pars{ C \bnabla \varPhi } = 0.
\end{array}\end{equation}
The macroscale Stokes flow equations have the same form as before \eqref{eq:stokes}:
\begin{equation}
\begin{array}{ccc}
\bnabla \cdot \bV = 0, &   
\Laplacian \bV - \bnabla P + \bnabla \varPhi \Laplacian \varPhi = \bzero.
\end{array}
\end{equation}

\subsection{Bulk-scale effective boundary conditions}
By integrating the equations across the Debye layer,
the effective boundary conditions at the particle surface are:
\subsubsection{Ion-exchanger} See \cite{yariv2010migration} for full derivation:
\begin{equation}
\begin{array}{ccc}
\label{eq:ionex_bnd}
0 = \varPhi + \log C, &
0 = \deriv{}{n} \pars{\varPhi - \log C}, &
\bV = 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C,
\end{array}\end{equation}
where $\zeta = \cV - \varPhi = \log (C / \gamma)$ 
is the zeta potential.

\subsubsection{Highly charged inert particle}
See \cite{schnitzer2012surface} for full derivation:
\begin{equation} \label{eq:ephor_bnd}
\begin{array}{ccc}
0 = {\deriv{C}{n} - \text{Du} \nabla^2_s \pars{\varPhi - \log C}}, & 
0 = {\deriv{C}{n} + C \deriv{\varPhi}{n}}, &
\bV = \zeta \cdot \bnabla_\mathcal{S} \pars{\varPhi - \log C} +
4 \log 2 \cdot \bnabla_\mathcal{S} \log C,
\end{array}\end{equation}
where $\zeta = \bar{\zeta} - \log C$ 
is the zeta potential and $\bar{\zeta} = 2 \log \sigma$.

\subsection{Steady state}
The stress tensor in the fluid is composed of Newtonian and Maxwel stresses:
\begin{equation}
\label{eq:tensor}
\tT = \bnabla \bV + (\bnabla \bV)^\dagger - P \tI
+ \bnabla \varPhi \bnabla \varPhi - \frac{1}{2} (\bnabla \varPhi \cdot \bnabla \varPhi). 
\end{equation} 
In steady state (constant particle drift velocity $\cU$), the total force acting on the particle 
(computed by the following surface integral) must vanish:
\begin{equation} \label{eq:zero_force}
 \bF(\beta, \cU) = \oint_\mathcal{S} \tT \cdot \bnhat dA = \bzero.
\end{equation}

The total force $\bF(\beta, \cU)$ is a function of the applied electric field $\beta$ and
the velocity $\cU$.
For any given electric field $\beta$, the force-free constraint above 
can be solved numerically for $\cU$ to yield the corresponding steady-state velocity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algorithm Description} \label{sec:algorithm}
The numerical solver first requires constructing a discrete 
approximation for the nonlinear PDE system,
for a given electric field $\beta$ and assumed velocity $\cU$. 
The resulting nonlinear discrete system is
solved numerically using Newton's Method, so the total force on the
particle can be computed. 
We seek the steady-state velocity $\cU$, 
which corresponds to zero total force, $F(\beta, \cU) = 0$,
which we solve by the standard secant method.

\subsection{Spherical Coordinates}
Because the system is axisymmetric, we employ
spherical coordinates $(r,\theta,\phi)$.
The scalar gradient and the divergence can be written as:
\begin{eqnarray}
\bnabla f = \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat, &
\bnabla \cdot \bF = \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}.
\end{eqnarray}
The scalar Laplacian can be written as:
\begin{eqnarray}
\Laplacian f = \bnabla \cdot (\bnabla f) = 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}.
\end{eqnarray}
The vector gradient and vector Laplacian can be written as:
\begin{eqnarray}
\bnabla \bF &=& \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat, \\
\bLaplacian \bF &=& 
\left(\Laplacian F_r - \frac{2F_r}{r^2} - 
\frac{2}{r^2 \sin\theta} \deriv{\left(F_\theta \sin\theta \right)}{\theta}\right)\brhat
+ \left(\Laplacian F_\theta - \frac{F_\theta}{r^2 \sin^2\theta} + 
\frac{2}{r^2}\deriv{F_r}{\theta}\right) \bthetahat.
\end{eqnarray}

Note that in the spherical coordinate system,
the vector Laplacian operator cannot be computed in a decoupled manner as in the Cartesian system,
since the unit vectors $\brhat, \bthetahat, \bphihat$ are functions of the coordinates
in a curvilinear coordinate system.

\subsection{Computational Grid}
A regular grid of size $n_r \times n_\theta$ is defined, with:
\begin{eqnarray} \label{eq:grid}
(r_i,\theta_j) \in [1, \infty) \times [0,\pi], & 
r_i = (1+\Delta_r)^i, &
 \theta_j = \Delta_\theta \cdot j,
\end{eqnarray}
where a logarithmic grid spacing has been chosen for $r$ and a uniform grid for $\theta$.
Note that $r_0 = 1$ and $R_{max} = (1+\Delta_r)^{n_r}$, where
$\Delta_r = \pars{R_{max}} ^ {1/n_r} - 1$ and $\Delta_\theta = {\pi/n_\theta}$.
Because the grid is finite, the choice 
$R_{max}$ must be taken large enough so as to have a negligible effect
on the solution.
This grid induces a disjoint subdivision of the domain 
$\Omega = [1, R_{max}] \times [0,\pi] = \bigcup_{ij}\Omega_{ij}$ into cells.
A specific cell $\Omega_{ij}$ and its center $(\bar{r}_i, \bar{\theta}_j)$ are defined by \eqref{eq:grid}:
\begin{eqnarray}
\Omega_{ij} = [r_{i-1}, r_{i}] \times [\theta_{j-1}, \theta_{j}], &
\bar{r}_i = (r_{i-1} + r_{i})/2, &
\bar{\theta}_j = (\theta_{j-1} + \theta_{j})/2.
\end{eqnarray}

Each discrete variable is located with respect of its cell as follows:
$\varPhi$, $C$ and $P$ are represented by their value at the center of each cell, 
using an all-centered grid, whereas $\bV$ is represented by its values at cell 
boundaries, using a staggered grid:
\begin{equation}
\begin{array}{ccccc}
\varPhi^h_{[i,j]} = \varPhi(\bar{r}_i, \bar{\theta}_j), &
C^h_{[i,j]} = C(\bar{r}_i, \bar{\theta}_j), &
P^h_{[i,j]} = P(\bar{r}_i, \bar{\theta}_j). &
V_r^h{}_{[i,j]} = V_r(r_i, \bar{\theta}_j), &
V_\theta^h{}_{[i,j]} = V_\theta(\bar{r}_i, {\theta}_j).
\end{array}
\end{equation}

\subsection{Operator Discretization} 
A finite-volume method with linear interpolation is used for flux discretization. 
Define the following discrete central difference operators:
\begin{eqnarray} \label{eq:disc_diff}
\cD_r(f^h){}_{\left[i+\half,j\right]} = \frac{f^h_{\left[i+1,j\right]} - f^h_{\left[i,j\right]}}
                       {r_{\left[i+1,j\right]} - r_{\left[i,j\right]}}, &
\cD_\theta(f^h){}_{\left[i,j+\half\right]} = \frac{f^h_{\left[i,j+1\right]} - f^h_{\left[i,j\right]}}
					   {\theta_{\left[i,j+1\right]} - \theta_{\left[i,j\right]}}.
\end{eqnarray}

Define the following linear interpolation operators:
\begin{eqnarray} \label{eq:disc_interp}
\begin{array}{rcl}
\cI_r(f^h)_{\left[i+\half,j\right]} &=& 
\frac{r_{\left[i+\half,j\right]} - r_{\left[i,j\right]}}
{r_{\left[i+1,j\right]} - r_{\left[i,j\right]}} 
f^h_{\left[i+1,j\right]} 
+ 
\frac{r_{\left[i+1,j\right]} - r_{\left[i+\half,j\right]}}
{r_{\left[i+1,j\right]} - r_{\left[i,j\right]}}
f^h_{\left[i,j\right]},
\\
\cI_\theta(f^h)_{\left[i,j+\half\right]} &=& 
\frac{r_{\left[i,j+\half\right]} - r_{\left[i,j\right]}}
{r_{\left[i,j+1\right]} - r_{\left[i,j\right]}}
  f^h_{\left[i,j+1\right]} + 
\frac{ r_{\left[i,j+1\right]} - r_{\left[i,j+\half\right]}}
{r_{\left[i,j+1\right]} - r_{\left[i,j\right]}}
  f^h_{\left[i,j\right]}.
\end{array}
\end{eqnarray}

If the total flux of a cell is equal to zero ($\bnabla \cdot \boldsymbol{f} = 0$), 
then using \eqref{eq:disc_diff}, we have
\begin{eqnarray}
\frac{1}{r^2} \cD_r\pars{f^h_r r^2} + 
\frac{1}{r \sin\theta} \cD_\theta\pars{f^h_\theta \sin\theta} =
\frac{\cD_r\pars{f^h_r \cdot r^2 \sin\theta} + 
\cD_\theta\pars{f^h_\theta \cdot r \sin\theta}}{r^2 \sin\theta} = 0. 
\end{eqnarray}

\subsection{System Equations} \label{sec:disc_equations}
Charge and salt fluxes \eqref{eq:bulk_flux} are discretized on grid cell boundaries 
(using an upwind scheme $\mathcal{U}$ for numerical stability at large cell Peclet number)  
using \eqref{eq:disc_diff} and \eqref{eq:disc_interp}:
\begin{eqnarray} \nonumber
I^h_r = -\cI_r(C^h) \cdot \cD_r(\varPhi^h), &
I^h_\theta = -\cI_\theta(C^h) \cdot \frac{\cD_\theta(\varPhi^h)}{r}, \\
J^h_r = -\cD_r(C^h) + \alpha V^h_r \cdot \mathcal{U}^{\bV^h}_r (C^h), &
J^h_\theta = -\frac{\cD_\theta(C^h)}{r} + \alpha V^h_\theta \cdot \mathcal{U}^{\bV^h}_\theta (C^h), 
\\ \nonumber
 \mathcal{U}^{\bV^h}_r(C^h)[i,j] = C^h\left[i-\frac{\sign(V^h_r)}{2}, j\right], &
 \mathcal{U}^{\bV^h}_\theta(C^h)[i,j] = C^h\left[i, j-\frac{\sign(V^h_\theta)}{2}\right]. 
\end{eqnarray}

Mass flux $\bV$ is discretized on grid cells boundaries, 
using the staggered velocity grid.

Force components ($\bF = -\bnabla P + \bLaplacian \bV + \bnabla \varPhi \Laplacian \varPhi$) 
are discretized on the staggered velocity grid, where
linear interpolation is used for the Coulomb force and for vector Laplacian components.
\begin{eqnarray}
\nonumber
F^h_r &=& -\cD_r(P^h) 
          + \cL(V^h_r) - \frac{2}{r^2} V^h_r 
		  - \frac{2}{r^2 \sin\theta} \cI_r(\cD_\theta (V^h_\theta \sin\theta))
          + \cD_r(\varPhi^h) \cdot \cI_r(\cL(\varPhi^h)), \\
F^h_\theta &=& -\frac{\cD_\theta(P^h)}{r} 
		  + \cL(V^h_\theta) - \frac{F^h_\theta}{r^2 \sin^2\theta} 
		  + \frac{2}{r^2} \cI_\theta(\cD_\theta(F^h_r))
		  + \frac{\cD_\theta(\varPhi^h)}{r} \cdot \cI_\theta(\cL(\varPhi^h)), \\
\nonumber
\cL(f^h) &=& \frac{1}{r^2}\cD_r\pars{\cD_r(f^h) r^2} + 
\frac{1}{r^2 \sin\theta} \cD_\theta\pars{\cD_\theta(f^h) \cdot \sin\theta}.
\end{eqnarray}

\subsection{Boundary Conditions} \label{sec:disc_boundary}
In order to discretize boundary condtions for $\varPhi$, $C$ and $\bV$, 
``ghost'' points are employed. 
The grid is extended to include points outside the domain interior,
and the variables at the ``ghost'' points are set to satisfy 
the discretized boundary conditions, following the analysis below.
Note that $P$ has no boundary conditions -- so the pressure variable is defined 
only in the domain interior.

Far away from the particle boundary \eqref{eq:bndcond_inf}, we have:
\begin{equation}
\begin{array}{cccc}
\cD_r \pars{\varPhi^h} = -\beta \cos\bar{\theta}, &
\cI_r \pars{C^h} = 1 &
V_r^h = -\cU \cos\theta, &
V_\theta^h = \cU \sin\theta.
\end{array}\end{equation}

For $\theta = 0$ and $\theta = \pi$, the ``ghost'' points are defined by axial 
symmetry considerations:
\begin{equation} 
\begin{array}{cccc}
\cD_\theta \pars{\varPhi^h} = 0, &
\cD_\theta \pars{C^h} = 0, &
\cD_\theta \pars{V_r^h} = 0, &
V_\theta^h = 0
\end{array}
\end{equation}

The specific boundary conditions at $r=1$
depend on the chemical properties of particle surface.

\subsubsection{Ion-exchanger} The boundary conditions \eqref{eq:ionex_bnd} 
are implemented by:
\begin{equation}
\begin{array}{rcl}
0 &=& \cI_r(\varPhi^h + \log C^h), \\
0 &=& \cD_r(\varPhi^h - \log C^h), \\
V^h_r &=& 0, \\
V^h_\theta &=& 4\log\pars{\frac{1}{2}\pars{1 + \exp\left\{\cI_\theta(\zeta^h)/2\right\}}} \cdot 
			\cD_\theta(\zeta^h), \\
  \zeta^h &=& - \log \gamma - \cI_r(\varPhi^h).
\end{array}
\end{equation}

\subsubsection{Highly charged inert particle} The boundary conditions \eqref{eq:ephor_bnd} 
are implemented by:
\begin{equation}
\begin{array}{rcl}
0 &=& \cD_r(C^h) - \text{Du} \cdot 
\pars{\cD_\theta \pars{\sin\theta \cdot \cD_\theta \cI_r\pars{\varPhi^h - \log C}}}/{\sin\theta}, \\
0 &=& \cD_r(C^h) + \cI_r(C^h) \cD_r(\varPhi^h), \\
V^h_r &=& 0, \\
V^h_\theta &=& \cI_\theta(\zeta^h) \cdot \cD_\theta(\cI_r \varPhi^h - \log \cI_r C^h) + 
4 \log 2 \cdot \cD_\theta(\log \cI_r C^h), \\
  \zeta^h &=& \bar{\zeta} - \log \cI_r C.
\end{array}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Solver Design}
An operator $\cO$ is defined by its input vector $\bx$ and its output 
$\by = \cO(\bx)$.
In order to solve the equation $\cO(\bx) = \bzero$, Newton's method is applied, 
given an initial solution $\bx_0$.
Each step requires the computation of the residual vector, $\br = \cO(\bx)$, and
the gradient matrix of the operator, $\bG = \bnabla \cO(\bx)$, 
given the current solution $\bx$.
Then, a sparse linear system $\bG \Delta \bx = \br$
needs to be solved to yield the update $\Delta \bx$.
When the method converges, $\bx_n \rightarrow \bx_\infty$, 
the asymptotic convergence is in general quadratic:
\begin{equation}
\|\bx_{n+1} - \bx_\infty\| \le k \|\bx_{n} - \bx_\infty\|^2.
\end{equation}
Thus, when the initial guess is close enough to the solution, the solver
will typically converge in very few steps.

\subsubsection{Operator Representation}
System variable $\bx$ is defined as the concatenation $\bx = [\,\varPhi, C, V_r, V_\theta, P\,]$.
Thus, each problem variable can be computed by applying an appropriate projection operator on $\bx$.
The system can be written as a nonlinear operator $\cO(\bx) = \bzero$ 
acting on the system variable ${\bx}$, 
where $\cO$ is the system equations operator, 
derived by concatenation of the system equations (described in Subsection \ref{sec:disc_equations})
and the boundary conditions operators (described in Subsection \ref{sec:disc_boundary}).
This way, the same numerical solver handles both the nonlinear system equation and nonlinear
boundary conditions.

\subsubsection{Steady-state}
Let $\beta$ be a given nondimensional electric field.
Assume that $\cU \ui$ is the drift velocity of the particle 
(given that the fluid is at rest far away from it).
After the convergence of the solver, once the residual norm is below
a prescribed threshold $\|\cO(\bx_n)\| < \epsilon$, 
the total force acting on the particle  
can be computed by integrating $\tT \cdot \bn$ over the particle surface.

>>>>

\documentclass[10pt]{article}
\hsize=5.5 true in
\textheight=8.4 true in
\topmargin 1in

@@

\title{Asymptotic Nonlinear Solution for 
Electrokinetic Flow Around an Ion-Exchange Spherical Particle}

% author and address
\author{Roman Zeyde \and Irad Yavneh}

@@

Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle. This structure, called
``Debye layer'', electrically screens the surface charge,
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle
size, an analytical asymptotic solution for the Debye layer
dynamics can be derived \cite{yariv2010asymptotic},
yielding ``effective'' boundary conditions near
the particle surface for the ``outer'' bulk region equations.

The variables of the electrokinetic flow equations are the electrostatic
potential $\varPhi$, fluid velocity $\bV$ and its pressure $P$, and
ionic concentration $C$.
The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.
The partial differential equations that describe the system dynamics
under an external electric field, are coupled and nonlinear, and
in general they have no analytic solution.

A closed-form linear asymptotic solution has been developed for
spherical particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration}. Once the electric field
becomes stronger, significant nonlinear phenomena emerge.
This interesting regime has not yet been explored extensively.

The goal of this work is to develop the next nonlinear terms
in the asymptotic solution of the ion-exchanger migration.
This solution can be used to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also for numerical solver results verification.

The derivation is performed as follows. The governing differential equations are
\begin{equation} \begin{array}{cccc}
\bnabla \cdot \pars{ C \bnabla \varPhi } = 0, &
\Laplacian C - \alpha \bV \cdot \bnabla C = 0, &
\Laplacian \bV - \bnabla P + \bnabla \varPhi \Laplacian \varPhi = \bzero, &
\bnabla \cdot \bV = 0.
\end{array}\end{equation}
The boundary conditions on the particle surface are
\begin{equation}
\begin{array}{cccc}
0 = \varPhi + \log C, &
0 = \deriv{}{n} \pars{\varPhi - \log C}, &
\bV = \zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C, &
\zeta = \log \frac{C}{\gamma}
\end{array}\end{equation}
Far away from the particle, the boundary conditions are
\begin{equation}\label{eq:bndcond_inf}\begin{array}{ccc}
\bV \rightarrow -\cU \ui, &
\bnabla \varPhi \rightarrow -\beta\ui, &
 C \rightarrow 1.
\end{array}\end{equation}

The system can be written as $\cO(\bx;\beta) = \bzero$.
For small $\beta$, the solution $\bx = [\varPhi; C; \bV; P]$ 
to the nonlinear system can be expanded in a Taylor series in $\beta$:
\begin{equation}
\bx(\beta) \approx \sum_n \bx_n \beta^n.
\end{equation}
The nonlinear terms
are expanded around $\beta = 0$:
\begin{equation}
\bzero = \cO(\bx) = \sum_i \cO_i(\bx_0, \ldots \bx_i) \beta^i.
\end{equation}
